T-reX: A tool to quantify waste and material footprints in current and future Life Cycle Assessment (LCA) databases

### 2024-04-01 latex_files/cover_letter.tex
update

We are pleased to submit our original research article titled ``T-reX: A Python Package to Quantify Supply Chain Flows of Waste and Material in LCA Databases'' for consideration by \textit{Resources, Conservation and Recycling}. This manuscript details the development and application of an innovative tool designed to improve the quantification of waste and material flows within Life Cycle Assessment (LCA) databases, an essential aspect of progressing towards sustainable resource management and circular economy objectives.

@@

The manuscript complies with the journal's requirements for an original research article, having a word count of [Insert Word Count], including [Insert Number of Figures] figures and [Insert Number of Tables] tables.

@@

\begin{enumerate}
    \item {Rafael Laurenti, IVL --- Swedish Environmental Research Institute, rafa@kth.se}
    \item {Valentina Bisinella,  DTU --- Technical University of Denmark, valenb@dtu.dk}
    \item {Reviewer 3 Name, Affiliation, Email Address}
\end{enumerate}
\vspace{1em}

>>>>

We are pleased to submit our original research article titled ``T-reX\: A Python Package to Quantify Supply Chain Flows of Waste and Material in LCA Databases'' for consideration by \textit{Resources, Conservation and Recycling}. This manuscript details the development and application of an innovative tool designed to improve the quantification of waste and material flows within Life Cycle Assessment (LCA) databases, an essential aspect of progressing towards sustainable resource management and circular economy objectives.

@@

The manuscript complies with the journal's requirements for an original research article, having a word count of 4969, including four figures and two tables.

@@

\begin{enumerate}
    \item {Rafael Laurenti, IVL --- Swedish Environmental Research Institute, rafa@kth.se}
    \item {Valentina Bisinella,  DTU --- Technical University of Denmark, valenb@dtu.dk}
    \item {Chris Mutel, Cauldron Solutions and brightway, cmutel@gmail.com}
\end{enumerate}
\vspace{1em}

### 2024-04-01 latex_files/introduction.tex
update

\cite{laurenti2023wastefootprint} developed a method to calculate the waste footprint of a product or service based on solving the demand vectors of the activities, also presenting simple measures to quantify waste hazardousness and `circularity'. In that study, it was shown that the waste footprint correlates well with other LCIA methods, particularly human health. The method presented, however, is limited in its scope and flexibility, is computationally intensive, difficult to use, is not easily reproducible, and suffers from errors due to double counting (as the authors themselves acknowledge). The `T-reX' Python software tool presented herein provides a more flexible, transparent, and user-friendly approach to quantifying waste flows in LCA\@. Moreover, the utility of T-reX is not limited to waste, it can be used to categorise and aggregate any technosphere exchange of interest (or customised grouping thereof), such as water, gas, and critical raw materials.

>>>>

~\cite{laurenti2023wastefootprint} developed a method to calculate the waste footprint of a product or service based on solving the demand vectors of the activities, also presenting simple measures to quantify waste hazardousness and `circularity'. In that study, it was shown that the waste footprint correlates well with other LCIA methods, particularly human health. The method presented, however, is limited in its scope and flexibility, is computationally intensive, difficult to use, is not easily reproducible, and suffers from errors due to double counting (as the authors themselves acknowledge). The `T-reX' Python software tool presented herein provides a more flexible, transparent, and user-friendly approach to quantifying waste flows in LCA\@. Moreover, the utility of T-reX is not limited to waste, it can be used to categorise and aggregate any technosphere exchange of interest (or customised grouping thereof), such as water, gas, and critical raw materials.

### 2024-04-01 latex_files/main.tex
auto-update
update

\linenumbers
% add something about the future stuffg
% SPOTLIGHTS (5 bullet points, max 120 characters each)
{\Large Spotlights}
\vspace{1em}
\begin{description}[style=nextline]
    \item[Bullet 1: Critical context and background information on the problem addressed] Tracking waste and material flows is essential for developing the circular economy and supply chain resilience
    \item[Bullet 2: A brief overview of the key finding of the study (or findings if necessary)] A tool was developed to quantify waste and material footprints of activities in life cycle assessment (LCA) databases
    \item[Bullet 3: The most radical, creative, disruptive or innovative aspect of the manuscript] Built to complement the Brightway LCA framework and ActivityBrowser, the tool is customisable and easy to use
    \item[Bullet 4: The significance of the results to the environment, economics, or society] The tool can be used to identify pre-consumer waste and material hotspots that are often hidden in supply chains
    \item[Bullet 5: Future vision or the most important implications for continued research] Improved data availability and quality would enable more detailed and accurate waste and material footprinting
\end{description}

\newpage
%% GRAPHICAL ABSTRACT
\begin{graphicalabstract}
    \cbox{to be improved :D}
    \includegraphics{grabs.jpg}
\end{graphicalabstract}

@@

    \begin{abstract}
        \cbox{Abstract word count: 155, Limit: 150}
        The quintessential principle of the `circular economy' is to minimise material consumption and waste generation.
        Thus, identifying and quantifying waste and material flows are of fundamental importance. Life Cycle Assessment (LCA) is a powerful method for this, given its capacity to pinpoint hotspots of environmental impact throughout the life cycle of activities, where the implementation of circular principles could be most effective.

@@

    % KEYWORDS (max 6)
    \begin{keyword}
        circular economy \sep waste \sep material \sep life cycle assessment \sep critical raw material \sep supply chain
    \end{keyword}

@@

% list of abbreviations
\begin{table}[H]
    \centering
    \caption{List of abbreviations}
    \label{tab:abbreviations}
    \begin{tabular}{ll}
        \toprule
        CF            & Carbon Footprint                                                    \\
        EF            & Ecological Footprint                                                \\
        CML           & Centrum voor Milieuwetenschappen (Centre for Environmental Science) \\
        CRM           & Critical Raw Material                                               \\
        CPC           & Cooperative Patent Classification                                   \\
        CSI           & Crustal Scarcity Indicator                                          \\
        CSP           & Crustal Scarcity Potential                                          \\
        EDIP          & Environmental Design of Industrial Products                         \\
        EEIOA         & Environmentally Extended Input-Output Analysis                      \\
        EOL           & End-of-Life                                                         \\
        FOEN          & Swiss Federal Office for the Environment                            \\
        IAM           & Integrated Assessment Model                                         \\
        IE            & Industrial Ecology                                                  \\
        LCA           & Life Cycle Assessment                                               \\
        LCI           & Life Cycle Inventory                                                \\
        LCIA          & Life Cycle Impact Assessment                                        \\
        LFP           & Lithium Iron Phosphate                                              \\
        Li-ion        & Lithium ion                                                         \\
        LiMn$_2$O$_4$ & Lithium Manganese Oxide                                             \\
        MF            & Material Footprint                                                  \\
        MFA           & Material Flow Analysis                                              \\
        NCA           & Nickel Cobalt Aluminum                                              \\
        NMC           & Nickel Manganese Cobalt                                             \\
        PWF           & Product Waste Footprint                                             \\
        RCP           & Representative Concentration Pathway                                \\
        ReCiPe        & A standard LCIA method set                                          \\
        SSPs          & Shared Socioeconomic Pathways                                       \\
        T-reX         & Tool for re-X in LCA                                                \\
        UBP           & Umweltbelastungspünkte (Environmental Impact Points)                \\
        UN            & United Nations                                                      \\
        WF            & Waste Footprint                                                     \\
        \bottomrule
    \end{tabular}
\end{table}

%% main text
\section{Introduction}
\label{sec:introduction}
\cbox{Section word count: 1700}
\input{introduction.tex}

\section{Methodology}
\label{sec:methodology}
\cbox{Section word count: 1600}
\input{methodology.tex}

\section{Results}
\cbox{Section word count: 1300}
\label{sec:results}
\input{results.tex}

\section{Discussion}
\label{sec:discussion}
\cbox{Section word count: 450}
\input{discussion.tex}

\section{Conclusions}
\label{sec:conclusions}
\cbox{Section word count: 150}

@@

The authors would like to acknowledge XXXX, XXXX, XXXX, XXX, who were part of the futures thinking small group session at the FutuRaM consortium meeting in Berlin in June 2023 in which the term `T-reX/T(ool)-reX' was coined to name the vision presented by Stewart Charles McDowall of a comprehensive opensource software framework for the study and development of secondary raw material systems. The authors hope to continue the development of T-reX toward this vision.

>>>>

\linenumbers{}

% SPOTLIGHTS (5 bullet points, max 120 characters each)

{\Large Spotlights}
\vspace{1em}
\begin{itemize}
    % Bullet 1: Critical context and background information on the problem addressed
    \item Tracking waste and material flows is essential for developing the circular economy and supply chain resilience
    % Bullet 2: A brief overview of the key finding of the study (or findings if necessary)
    \item A tool was developed to quantify waste and material footprints of activities in life cycle assessment (LCA) databases
    % Bullet 3: The most radical, creative, disruptive or innovative aspect of the manuscript
    \item Built to complement the Brightway LCA framework and ActivityBrowser, the tool is customisable and easy to use
    % Bullet 4: The significance of the results to the environment, economics, or society
    \item The tool can be used to identify pre-consumer waste and material hotspots that are often hidden in supply chains
    % Bullet 5: Future vision or the most important implications for continued research
    \item Improved data availability and quality would enable more detailed and accurate waste and material footprinting
\end{itemize}

\newpage
%% GRAPHICAL ABSTRACT
\begin{graphicalabstract}
    \includegraphics{grabs.jpg}
\end{graphicalabstract}

@@

    \begin{abstract}
        % \cbox{Abstract word count: 155, Limit: 150}
        The quintessential principle of the `circular economy' is to minimise material consumption and waste generation.
        Thus, identifying and quantifying waste and material flows are of fundamental importance. Life Cycle Assessment (LCA) is a powerful method for this, given its capacity to pinpoint hotspots of environmental impact throughout the life cycle of activities, where the implementation of circular principles could be most effective.

@@

    % KEYWORDS (max 6)
    \begin{keyword}
        circular economy \sep{}waste \sep{}material \sep{}life cycle assessment \sep{}critical raw material \sep{}supply chain
    \end{keyword}

@@

% list of abbreviations
\begin{table}[H]
    \centering
    \caption{List of abbreviations}\label{tab:abbreviations}
    \begin{tabular}{ll}
        \toprule
        CF            & Carbon Footprint                                                    \\
        EF            & Ecological Footprint                                                \\
        CML           & Centrum voor Milieuwetenschappen (Centre for Environmental Science) \\
        CRM           & Critical Raw Material                                               \\
        CPC           & Cooperative Patent Classification                                   \\
        CSI           & Crustal Scarcity Indicator                                          \\
        CSP           & Crustal Scarcity Potential                                          \\
        EDIP          & Environmental Design of Industrial Products                         \\
        EEIOA         & Environmentally Extended Input-Output Analysis                      \\
        EOL           & End-of-Life                                                         \\
        FOEN          & Swiss Federal Office for the Environment                            \\
        IAM           & Integrated Assessment Model                                         \\
        IE            & Industrial Ecology                                                  \\
        LCA           & Life Cycle Assessment                                               \\
        LCI           & Life Cycle Inventory                                                \\
        LCIA          & Life Cycle Impact Assessment                                        \\
        LFP           & Lithium Iron Phosphate                                              \\
        Li-ion        & Lithium ion                                                         \\
        LiMn\(_2\)O\(_4\) & Lithium Manganese Oxide                                             \\
        MF            & Material Footprint                                                  \\
        MFA           & Material Flow Analysis                                              \\
        NCA           & Nickel Cobalt Aluminum                                              \\
        NMC           & Nickel Manganese Cobalt                                             \\
        PWF           & Product Waste Footprint                                             \\
        RCP           & Representative Concentration Pathway                                \\
        ReCiPe        & A standard LCIA method set                                          \\
        SSPs          & Shared Socioeconomic Pathways                                       \\
        T-reX         & Tool for re-X in LCA                                                \\
        UBP           & Umweltbelastungspünkte (Environmental Impact Points)                \\
        UN            & United Nations                                                      \\
        WF            & Waste Footprint                                                     \\
        \bottomrule
    \end{tabular}
\end{table}

%% main text
\section{Introduction}\label{sec:introduction}
%\cbox{Section word count: 2000}
\input{introduction.tex}

\section{Methodology}\label{sec:methodology}
%\cbox{Section word count: 1700}
\input{methodology.tex}

\section{Results}\label{sec:results}
% \cbox{Section word count: 1200}
\input{results.tex}

\section{Discussion}\label{sec:discussion}
% \cbox{Section word count: 450}
\input{discussion.tex}

\section{Conclusions}\label{sec:conclusions}
% \cbox{Section word count: 150}

@@

The authors would like to acknowledge those who were part of the futures thinking small group session at the FutuRaM consortium meeting in Berlin in June 2023 in which the term `T-reX/T(ool)-reX' was coined to name the vision presented by Stewart Charles McDowall of a comprehensive opensource software framework for the study and development of secondary raw material systems. The authors hope to continue the development of T-reX toward this vision.

### 2024-04-01 latex_files/methodology.tex
update

This section is divided into two parts. In \autoref{sec:method-T-reX_tool}, we describe the T-reX, and in \autoref{sec:method-case_study}, we describe the methodology used to calculate the waste and material inventory footprints in a case study of five Li-ion batteries.

\subsection{T-reX}\label{sec:method-T-reX_tool}

@@

Developed in the Python programming language, T-reX extends the \texttt{brightway LCA framework}, utilising the components \texttt{bw2data}, \texttt{bw2calc}, and \texttt{bw2io}~\citep{mutel2017brightway}. Additionally, the \texttt{wurst} package---which can expand entire databases into a list of exchanges---is used to facilitate database searching and data transformation at the exchange level~\citep{mutel2017wurst}. Integration with the \texttt{premise} package~\citep{sacchi2022premise}---which integrates the projections of integrated assessment models (IAMs) into current LCA databases---enables the user to easily create and manipulate prospective LCA databases. T-reX is also compatible with \texttt{ActivityBrowser}~\citep{steubing2020activitybrowser}---an open-source graphical user interface for LCA---after running T-reX, the manipulated databases and the 'pseudo- LCIA' methods created by T-reX can be used in the accustomed way. T-reX is installable via the Python Package Index (PyPI)~\citep{mcdowall2023T-reXpipy} and is open source under the CC-0 licence. The full source code for T-reX is indexed on Zenodo~\citep{mcdowall2023T-reXzenodo} and under further development in the GitHub repository~\citep{mcdowall2024T-reXgithub}. T-reX is designed to be used with ecoinvent databases~\citep{ecoinvent2016version3}, but could be adapted to other databases by changing the search criteria. Currently, it has been tested with all available system models of ecoinvent versions 3.5--3.10.

@@

T-reX, in contrast, uses accounting for material demand (and waste generation) based on exchanges solely within the technosphere. This offers a different perspective, allowing for the estimation of overall supply chain material demands that consider the entire life cycle of an activity, including non-direct impacts on the market such as co-production of other materials. Consider a demand for an activity containing a metal, for example; while the existing material use methods allow one to calculate the total mass of that metal that is extracted from the environment, T-reX can provide insight into the broader supply chain impacts of the demand for this metal. If the production of other materials is attributed to the production of this metal, these would appear as negative material demands in the T-reX results---supply chain pressure for one material can result in lessening of supply chain pressure for another. In the results of the Li-ion battery case study in \autoref{sec:results-case_study}, we will see that this is indeed the case for nickel demand, which, in the final inventory because of such co-production, is counter-intuitively negative (due to co-production and substitution in the LCA system models)despite the presence of nickel in the final products.

@@

This workflow creates a copy of the original \texttt{brightway} project containing the original biosphere database, a T-reX `pseudo-biosphere' database along with one or more manipulated technosphere database(s) that can be used to calculate the waste and material inventory footprints of activities in the same way as standard LCIA calculations.

@@

\begin{figure}[H]
    \centering
    \caption{Workflow of T-reX. Application of T-reX to one or more LCA databases creates a brightway project with customised methods and activities that can be used to calculate waste and material inventory footprints of the activity's supply chain.}
    \includegraphics[width=0.5\linewidth]{figures/T-reX_method.pdf}
    \label{fig:methods-flowchart}
\end{figure}

\subsection{Case study methodology}\label{sec:method-case_study}

>>>>

This section is divided into two parts. In \autoref{sec:method-TreX}, we describe the T-reX program, and in \autoref{sec:method-case-study}, we describe the methodology used to calculate the waste and material inventory footprints in a case study of five Li-ion batteries.

\subsection{T-reX}\label{sec:method-TreX}

@@

Developed in the Python programming language, T-reX extends the \texttt{brightway LCA framework}, utilising the components \texttt{bw2data}, \texttt{bw2calc}, and \texttt{bw2io}~\citep{mutel2017brightway}. Additionally, the \texttt{wurst} package---which can expand entire databases into a list of exchanges---is used to facilitate database searching and data transformation at the exchange level~\citep{mutel2017wurst}. Integration with the \texttt{premise} package~\citep{sacchi2022premise}---which integrates the projections of integrated assessment models (IAMs) into current LCA databases---enables the user to easily create and manipulate prospective LCA databases. T-reX is also compatible with \texttt{ActivityBrowser}~\citep{steubing2020activitybrowser}---an open-source graphical user interface for LCA---after running T-reX, the manipulated databases and the `pseudo- LCIA' methods created by T-reX can be used in the accustomed way. T-reX is installable via the Python Package Index (PyPI)~\citep{mcdowall2023T-reXpipy} and is open source under the CC-0 licence. The full source code for T-reX is indexed on Zenodo~\citep{mcdowall2023T-reXzenodo} and under further development in the GitHub repository~\citep{mcdowall2024T-reXgithub}. T-reX is designed to be used with ecoinvent databases~\citep{ecoinvent2016version3}, but could be adapted to other databases by changing the search criteria. Currently, it has been tested with all available system models of ecoinvent versions 3.5--3.10.

@@

T-reX, in contrast, uses accounting for material demand (and waste generation) based on exchanges solely within the technosphere. This offers a different perspective, allowing for the estimation of overall supply chain material demands that consider the entire life cycle of an activity, including non-direct impacts on the market such as co-production of other materials. Consider a demand for an activity containing a metal, for example; while the existing material use methods allow one to calculate the total mass of that metal that is extracted from the environment, T-reX can provide insight into the broader supply chain impacts of the demand for this metal. If the production of other materials is attributed to the production of this metal, these would appear as negative material demands in the T-reX results---supply chain pressure for one material can result in lessening of supply chain pressure for another. In the results of the Li-ion battery case study in \autoref{sec:results-case_study}, we will see that this is indeed the case for nickel demand, which, in the final inventory because of such co-production, is counter-intuitively negative (due to co-production and substitution in the LCA system models) despite the presence of nickel in the final products.

@@

This workflow creates a copy of the original \texttt{brightway} project containing the original biosphere database, a T-reX `pseudo-biosphere' database along with one or more manipulated technosphere databases that can be used to calculate the waste and material inventory footprints of activities in the same way as standard LCIA calculations.

@@

\begin{figure}[H]
    \centering
    \caption{Workflow of T-reX. Application of T-reX to one or more LCA databases creates a brightway project with customised methods and activities that can be used to calculate waste and material inventory footprints of the activity's supply chain.}
    \includegraphics[width=0.5\linewidth]{figures/T-reX_method.pdf}\label{fig:methods-flowchart}
\end{figure}

\subsection{Case study methodology}\label{sec:method-case-study}

### 2024-04-01 latex_files/results.tex
update

As described in~\autoref{sec:method-case_study}, this case study calculated the waste and material footprints (with a variety of other indicators) for the unaltered inventories of five Li-ion batteries with the functional unit being 1~kg of the battery at the global market. The purpose of this simple case study was to test, verify, and demonstrate the functionality and limitations of T-reX. This section includes some highlights of the results with full results included in the supplementary material\autoref{sec:supplementary}. Because the `pseudo LCIA' T-reX methods are integrated into the brightway project as if they were LCIA methods, the LCA calculations can be performed in the customary way. In the supplementary material, there are screenshots of selected results obtained using the \texttt{ActivityBrowser} software, including contribution analysis and a Sankey diagram that disaggregates the final footprint result over the activities in the supply chain.

>>>>

As described in~\autoref{sec:method-case-study}, this case study calculated the waste and material footprints (with a variety of other indicators) for the unaltered inventories of five Li-ion batteries with the functional unit being 1~kg of the battery at the global market. The purpose of this simple case study was to test, verify, and demonstrate the functionality and limitations of T-reX. This section includes some highlights of the results with full results included in the supplementary material\autoref{sec:supplementary}. Because the `pseudo LCIA' T-reX methods are integrated into the brightway project as if they were LCIA methods, the LCA calculations can be performed in the customary way. In the supplementary material, there are screenshots of selected results obtained using the \texttt{ActivityBrowser} software, including contribution analysis and a Sankey diagram that disaggregates the final footprint result over the activities in the supply chain.
